Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-04-30 PAPER.tex
Wed Apr 25 12:26:14 EDT 2018
updated fig 22 sensitivity comparison to reflect A15 analysis, and edited text to highlight where changes have been made to this analysis

\newcommand{\invC}{\ensuremath{C^{-1}}}
\newcommand{\dC}[1]{\ensuremath{C_{,#1}}}
\DeclareMathOperator{\Tr}{Tr}
\newcommand{\half}{\ensuremath{\frac{1}{2}}}
\newcommand{\PDeriv}[2]{\ensuremath{\frac{\partial #1}{\partial #2}}}

@@

% Put all of the (O)QE / ICW theory here, rather than sprinkled throughout this section.

% Some things we never quite say as explicitly as I'd like:
% - "foreground suppression" happens because the foregrounds have a different covariance structure (Figure 4), and we 
% hope we pick this out by picking big eigenvalues
% - additive bias is a thing

\subsubsection{The Quadratic Estimator Method}

@@

@@

%In practice, we do not have perfect models for $\textbf{C}$, and it is the use of an estimated covariance $\widehat{\textbf{C}}$ instead of the true covariance $\textbf{C}$ that can lead to loss.

@@

% ---

\subsubsection{Toy Model: Empirical Inverse Covariance Weighting}
\label{sec:toymodel}

We will now build intuition into how empirically estimated inverse covariance weighting can give rise to signal loss through the use of a toy model.  We construct a simple dataset that contains visibility data with $100$ time 
integrations and $20$ frequency channels. This model represents realistic dimensions of about an hour of PAPER data which 
might be used for a power spectrum analysis. For PAPER-64 (both the \citetalias{ali_et_al2015} analysis and our new analysis) we use $\sim8$ hours of data (with channel 
widths of $0.5$ MHz and integration times of $43$ seconds), but here we scale it down with no loss of 
generality.

@@

% While a later section does show that the large eigenvalue modes converge to the eigenvectors quicker, it is _not_ true that modes with big eigenvalues are downweighted more.  It is the modes which have a structure different from dC/dp_alpha.  Our regularization actually forces both eigenvalues and eigenmodes to be like the "signal" part of the covariance.  Thus, I don't think the next two paragraphs are helping the explanation of what is going on.

%To begin to understand the lossy behavior of this result, we can closely study our estimated covariance eigenspectrum, shown in Figure \ref{fig:toy_sigloss2}. Since it is estimated from our data, its eigenspectrum differs from the eigenspectrum of the true covariance $\textbf{C}$, and this difference has important consequences on our result.

%An eigenspectrum ranks the eigenvalues of a matrix from highest to lowest and can be thought of as a spectrum of weights that are given to each spectral mode in the data. In other words, the eigenvalues encode the strength of different shapes in the dataset. The eigenspectrum of the identity matrix $ \textbf{I}$ is flat (all $1$'s) because it gives equal weighting to all modes. In our application, covariance matrices tend to have sloped eigenspectra, meaning that modes are given different weights in QE power spectrum estimation. The modes with the highest eigenvalues are down-weighted the most.

@@

Because $\widehat{\textbf{C}}$ is estimated from the data, its eigenvectors and eigenvalues are strongly coupled to the particular data realization that was used to compute it.  In this case, the eigenmode of $\widehat{\textbf{C}}$ with the largest eigenvalue describes the sinusoid foreground mode in the toy model (the peak in Figure \ref{fig:toy_sigloss3}).  In Figure \ref{fig:toy_sigloss12} we show the estimated covariances of our toy model datasets along with $\widehat{\textbf{C}}^{-1}$ weighted data. The foreground sinusoid is clearly visible in $\widehat{\textbf{C}}_{\rm FG}$.

%In general, 
We have made the assumption that the highest-valued eigenmodes of $\widehat{\textbf{C}}$ will typically (but not always) describe bright foregrounds, though we have not checked that indeed the eigenvectors do describe correlations between frequencies.
% --- the most prominent shapes in a dataset. 
%For these FG-dominating modes, where foregrounds outshine the EoR signal, 
If such modes do describe foregrounds, then down-weighting is beneficial. In some sense, we desire signal loss in this regime, if by `signal' we mean `foregrounds.'  In this case it is beneficial for these eigenmodes to be coupled to the data in a way that produces loss. For our toy model, the successful suppression of the foreground mode is demonstrated in Figure \ref{fig:toy_sigloss3} by the missing foreground peak in the weighted power spectrum estimate (green).

@@

analysis in this paper, we only use one type of baseline (PAPER's 30\,m East/West baselines), hence $N_{\rm seps}=1$. The updated limits in Kolopanis et al. (\textit{submitted})
use three separation types.

@@

the total number of $30$\,m East/West baselines used in the analysis. However, using the total number of baselines ($N_{\rm bls\_total}$) neglects 
the fact that we average baselines into groups for computational speed-up when cross-multiplying data. Our revised estimate for the parameter is:

@@

\noindent where, as with the \citetalias{ali_et_al2015} analysis, $N_{\rm gps} = 5$. Each baseline group averages down linearly as the number of baselines 

@@

N_{\rm gps}}{2}}$\Big). For the revised PAPER-64 analysis with only one baseline separation type, this becomes $N_{\rm bls} \sim 32$ instead 
of $51$. 

@@

Our revised sensitivity estimate for PAPER-64 is shown in comparison with that of \citetalias{ali_et_al2015} in Figure \ref{fig:sense_check}. 
Together, the revised parameters yield a decrease in sensitivity (higher noise floor) by a factor of $\sim5$ in mK$^{2}$.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{plots/sense_check.pdf}
	\caption{An updated prediction for the thermal noise level of PAPER-64 data (black) is shown in comparison to previously 
published sensitivity limits (gray). Both sensitivity analyses plotted assume only one baseline type (an additional factor of $
\sqrt{3}$ for three baseline types is needed to match \citetalias{ali_et_al2015} exactly). Major factors that contribute to the discrepancy are $
\Omega_{\rm eff}$, $N_{\rm days}$ and $N_{\rm bls}$, as in Equation \eqref{eq:sense} and described in Section \ref{sec:PSSense}, which when combined decreases our 
sensitivity (higher noise floor) by a factor of $\sim5$ in mK$^{2}$.}
	\label{fig:sense_check}
\end{figure}

@@

Power spectrum results for the noise simulation, which uses our full power spectrum pipeline, are shown in Figure 
\ref{fig:ps_noise}, where the black and gray points represent pre-signal loss estimated positive and negative power spectrum values, respectively (with 
$2\sigma$ error bars and weighting matrix $\widehat{\textbf{C}}_{\rm eff}$), the dashed black line represents the post-signal loss $2\sigma$ upper limit on the EoR signal, the dashed blue line represents the $2\sigma$ uniform-weighted power 
spectrum limit, and the solid green line denotes our $2\sigma$ theoretical noise prediction as calculated by Equation 
\eqref{eq:sense}. All three show good agreement to within a factor of two, which we believe is reasonable given the approximations we make in computing the analytic expression for thermal noise.

@@

\begin{figure}
	\centering
	\includegraphics[trim={0.3cm 0cm 0.3cm 0.3cm},width=\columnwidth]{plots/noise_errors.pdf}
	\caption{$2\sigma$ power spectrum errors for a noise simulation (computed via Equation \eqref{eq:noise} using PAPER-64 observing parameters) using two different bootstrapping 
methods. The noise is fringe-rate filtered and a weighting matrix of $\textbf{I}$ (uniform-weighted) is used in order to disentangle the 
effects of bootstrapping from signal loss. The bootstrapping method used in \citetalias{ali_et_al2015} is shown in gray, where bootstrapping occurs along both the baseline and time axes. This under-estimates errors by sampling more values than independent ones in the dataset (fringe-rate filtering reduces the number of independent samples along time). We use the method illustrated by the black curve in our updated analysis, where bootstrapping only occurs along the baseline axis. While this does not match perfectly with the $2\sigma$ analytic prediction for noise (green), we find good agreement between them to within a factor of two (which we believe is reasonable given the approximations we make in computing the analytic).}
	\label{fig:data_errors}
\end{figure}

@@

\color{blue}

\appendix
\section{A toy model for inverse covariance weighting}
\label{sec:icw_appendix}

We illustrate the way in which the quadratic estimator, Equation \ref{eq:QuadraticEstimator}, is intended to estimate the power spectrum of EoR in the presence of contamination by other terms by considering the simpler, but similar, case of estimating the variance of data.  We have specifically chosen a toy model where the data covariance is diagonal, as we expect the EoR signal to be.  We assume we have $N$ data points $\Delta_i$ which are the sum of a desired signal $\sigma_i$ and some contaminant $\phi_i$
\begin{equation}
\Delta_i = \sigma_i + \phi_i
\end{equation}
with 
\begin{equation}
\langle \sigma_i \rangle = 0; \; \langle \sigma^2_i \rangle = s; \; {\rm and}~\langle \sigma \sigma^T \rangle = s I_{N \times N} \equiv S
\end{equation}
and we wish to estimate $s$.  The contaminant in this case has a similar structure for its covariance, and is assumed uncorrelated with the signal
\begin{equation}
\langle \phi_i \rangle = 0; \; 
\langle \phi^2_i \rangle = f; \; 
\langle \phi \phi^T \rangle = f I_{N \times N} \equiv F; {\rm and}~ 
\langle \sigma_i \phi_j \rangle = 0
\end{equation}
%That is, we have the sum of $N$ independent samples of two random variables and we wish to estimate the variance of one of the random variables in the sum; in this case, $s$.
The estimator for $s$ using the inverse covariance weighting prescription is
\begin{equation}
\label{eq:CaseOneICW}
\hat{s} = \half {\cal F}^{-1} \left( \Delta^T \invC \PDeriv{C}{s}  \invC \Delta \right) 
= \frac{\Delta^T \Delta}{N}
\end{equation}
%and plugging in the relevant quantities gives
%\begin{eqnarray}
%\hat{s} & = & \half \frac{2}{N} (s+g)^2 \left( \Delta^T \frac{1}{s+g} I \frac{1}{s+g} \Delta \right) \\
%& = & \frac{\Delta^T \Delta}{N}
%\end{eqnarray}
We note that
\begin{equation}
\langle \hat{s} \rangle = s + g
\end{equation}
In this case, the quadratic estimator returns an entirely reasonable result, but, 
{\it since the covariance structure of the contaminant is identical to the signal}, it is only possible to achieve an unbiased estimate by using some additional information to estimate the additive bias and subtract it.

Now consider a case, chosen to be very similar to the toy model in \ref{sec:},  in which the data have an additive contaminant given by
\begin{equation}
\Delta_i = \sigma_i + \phi m_i
\end{equation}
where the properties of $\sigma_i$ are as before,
%\begin{eqnarray}
%\langle \sigma_i \rangle & = & 0 \\
%\langle \sigma^2_i \rangle & = & s \\
% \langle \sigma \sigma^T \rangle & = & s I_{N \times N} \equiv S
%\end{eqnarray}
but now $\phi$ is a random variable and $m_i$ is a fixed function of $i$ with
\begin{equation}
\langle \phi \rangle = 0; \;
\langle \phi^2 \rangle = f; \; 
 \langle \phi \phi^T \rangle = f m m^T \equiv F; \;
 m^T m = 1
\end{equation}
Here $m$ represents a mode which is correlated across many data points, with correlation given by $m m^T$, and, and the amplitude of its realization is $\phi$.

\color{black}

>>>>

@@

@@

\subsubsection{Toy Model: Inverse Covariance Weighting}
\label{sec:toymodel}

@@

In practice, we do not have perfect models for $\textbf{C}$, and it is the use of an estimated covariance $\widehat{\textbf{C}}$ instead of the true covariance $\textbf{C}$ that can lead to loss. We will now build intuition into how empirically estimated inverse covariance weighting can give rise to signal loss through the use of a toy model.

We construct a simple dataset that contains visibility data with $100$ time 
integrations and $20$ frequency channels. This model represents realistic dimensions of about an hour of PAPER data which 
might be used for a power spectrum analysis. For PAPER-64 (both the \citetalias{ali_et_al2015} analysis and our new analysis) we use $\sim8$ hours of data (with channel 
widths of $0.5$ MHz and integration times of $43$ seconds), but here we scale it down with no loss of 
generality.

@@

@@

To begin to understand the lossy behavior of this result, we can closely study our estimated covariance eigenspectrum, shown in Figure \ref{fig:toy_sigloss2}. Since it is estimated from our data, its eigenspectrum differs from the eigenspectrum of the true covariance $\textbf{C}$, and this difference has important consequences on our result.

An eigenspectrum ranks the eigenvalues of a matrix from 
highest to lowest and can be thought of as a spectrum of weights that are given to each spectral mode in the data. In other 
words, the eigenvalues encode the strength of different shapes in the dataset. The eigenspectrum of the identity matrix $
\textbf{I}$ is flat (all $1$'s) because it gives equal weighting to all modes. In our application, covariance matrices tend to have sloped eigenspectra, meaning that modes are given different weights in QE power spectrum estimation. The modes with the highest eigenvalues are 
down-weighted the most.

@@

Because $\widehat{\textbf{C}}$ is estimated from the data, its eigenvectors and eigenvalues are strongly coupled to the particular data realization that was used to compute it. For example, the highest-valued eigenmode of $\widehat{\textbf{C}}$ (highest eigenvalue) describes the sinusoid foreground mode in the toy model (the peak in Figure \ref{fig:toy_sigloss3}). In Figure \ref{fig:toy_sigloss12} we show the estimated covariances of our toy model datasets along with $\widehat{\textbf{C}}^{-1}$ weighted data. The foreground sinusoid is clearly visible in $\widehat{\textbf{C}}_{\rm FG}$.

In general, the highest-valued eigenmodes of $\widehat{\textbf{C}}$ typically (but not always) describe bright foregrounds --- the most prominent shapes in a dataset. For these FG-dominating modes, where foregrounds outshine the EoR signal, down-weighting is beneficial. In some sense, we desire signal loss in this regime, if by `signal' we mean `foregrounds.' In this case it is beneficial for these eigenmodes to be coupled to the data in a way that produces loss. For our toy model, the successful suppression of the foreground mode is demonstrated in Figure \ref{fig:toy_sigloss3} by the missing foreground peak in the weighted power spectrum estimate (green).

@@

analysis in this paper, we only use one type of baseline (PAPER's 30\,m East/West baselines). However, both the updated limits in Kolopanis et al. (\textit{submitted}) and the sensitivity prediction in Figure \ref{fig:sense_check} use three separation types ($N_{\rm seps}=3$) to match \citetalias{ali_et_al2015}.

@@

the total number of $30$\,m East/West baselines used in the analysis. However, using the total number of baselines ($N_{\rm bls\_total} = 51$) neglects 
the fact that the \citetalias{ali_et_al2015} analysis averages baselines into groups for computational speed-up when cross-multiplying data. Our revised estimate for the parameter is:

@@

\noindent where, in the \citetalias{ali_et_al2015} analysis, $N_{\rm gps} = 5$. Each baseline group averages down linearly as the number of baselines 

@@

N_{\rm gps}}{2}}$\Big). A revised \citetalias{ali_et_al2015} analysis should therefore use $N_{\rm bls} \sim 32$ instead of $51$, and this change is taken into account in Figure \ref{fig:sense_check}. However, the analysis in this paper and in Kolopanis et al. (\textit{submitted}) does not average baselines into groups ($N_{\rm gps} = 1$), and hence $N_{\rm bls} = 51$ for the revised PAPER-64 analysis.

@@

Our revised sensitivity estimate for the \citetalias{ali_et_al2015} analysis of PAPER-64 is shown in Figure \ref{fig:sense_check}. 
Together, the revised parameters yield a decrease in sensitivity (higher noise floor) by a factor of $\sim7$ in mK$^{2}$.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{plots/sense_check.pdf}
	\caption{An updated prediction for the thermal noise level of PAPER-64 data (black) is shown in comparison to previously 
published sensitivity limits (gray), both computed for the analysis methods used in \citetalias{ali_et_al2015}. Major factors that contribute to the discrepancy are $
\Omega_{\rm eff}$, $N_{\rm days}$ and $N_{\rm bls}$, as in Equation \eqref{eq:sense} and described in Section \ref{sec:PSSense}, which when combined decreases our 
sensitivity (higher noise floor) by a factor of $\sim7$ in mK$^{2}$.}
	\label{fig:sense_check}
\end{figure}

@@

Power spectrum results for the noise simulation, which uses our full power spectrum pipeline, are shown in Figure 
\ref{fig:ps_noise}, where the black and gray points represent pre-signal loss estimated positive and negative power spectrum values, respectively (with 
$2\sigma$ error bars and weighting matrix $\widehat{\textbf{C}}_{\rm eff}$), the dashed black line represents the post-signal loss $2\sigma$ upper limit on the EoR signal, the dashed blue line represents the $2\sigma$ uniform-weighted power 
spectrum limit, and the solid green line denotes our $2\sigma$ theoretical noise prediction as calculated by Equation 
\eqref{eq:sense}. All three show good agreement.

@@

\begin{figure}
	\centering
	\includegraphics[trim={0.3cm 0cm 0.3cm 0.3cm},width=\columnwidth]{plots/noise_errors.pdf}
	\caption{$2\sigma$ power spectrum upper limits for a noise simulation (computed via Equation \eqref{eq:noise} using PAPER-64 observing parameters) using two different bootstrapping 
methods. The noise is fringe-rate filtered and a weighting matrix of $\textbf{I}$ (uniform-weighted) is used in order to disentangle the 
effects of bootstrapping from signal loss. The bootstrapping method used in \citetalias{ali_et_al2015} is shown in gray, where bootstrapping occurs along both the baseline and time axes. This under-estimates errors by sampling more values than independent ones in the dataset (fringe-rate filtering reduces the number of independent samples along time). We use the method illustrated by the black curve in our updated analysis, where bootstrapping only occurs along the baseline axis. We find that these revised limits agree with the $2\sigma$ analytic prediction for noise (green).}
	\label{fig:data_errors}
\end{figure}

@@
